---
title: 'Advanced Statistical Modelling: Ridge Regression'
author: "Ricard Meyerhofer & Joel Cantero"
date: "29/10/2019"
output:
  pdf_document:
    latex_engine: xelatex
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library("ggplot2")
library("car")
library("caret")
library("dplyr")
library("GGally")
library("kableExtra")
library("glmnet")
library("plotmo")
library("data.table")
set.seed(42)
```

# Choosing the penalization parameter $\lambda$
The objective of this exercise is to implement Ridge Regression with two different approaches: $MSPE_{val}(\lambda)$ and $MSPE_{k-CV}(\lambda)$. In both cases we are going to take as input data the following:
\begin{itemize}
\item Matrix x and vector y corresponding to the training sample.
\item Matrix $x_{val}$ and vector $y_{val}$ corresponding to the validation set.
\item Vector $lambda.v$ of candidate values for $\lambda$.
\end{itemize}
We are going to output for each element $\lambda$ in $lambda.v$ and the value of the $MSPE_{val}(\lambda)$ / $MSPE_{k-CV}(\lambda)$. Furthermore, we are going to plot these values against $log(1+ \lambda)-1$.

Once we have build these two functions, we are going to use the prostate data used in class. We are going to choose a $\lambda$ according to:
\begin{itemize}
\item Behaviour in the validation set (30 validations not included in the training sample)
\item 5-fold, 10-fold cross-validation.
\item Compare our results with those obtained with leave-one-out and generalized cross-validation.
\end{itemize}

## Ridge regression based on $MSPE_{val}(\lambda)$ 
In order to choose the penalization parameter $\lambda$, we are going to write a function implementing the ridge regression penalization parameter $\lambda$ choice based on the minimization of the $MSPE_{val}(\lambda)$.

```{r lambda_function}
ridge <- function(x, y, x.val, y.val, lambda.v) {
  result <- data.frame(lambda=lambda.v, mspe=rep(0,length(lambda.v)), 
                       df=rep(0,length(lambda.v)))
  x.svd <- svd(x)
  for(i in 1:length(lambda.v)) {
    lambda <- lambda.v[i]
    d_inv <- diag(1/(x.svd$d*x.svd$d - lambda))
    xx_inv <- t( solve( t(x) %*% x + lambda*diag(1,ncol(x)) ))
    beta <- xx_inv %*% t(x) %*% y
    y.hat <- x.val %*% beta
    mspe <- sum((y.val - y.hat)^2) / length(y.val)
    df <- sum(x.svd$d^2 / (x.svd$d^2 +lambda))
    result$mspe[i] <- mspe
    result$df[i] <- df
  }
  plot(mspe~log(1+lambda), result, col=2)
  lambda.min <- result$lambda[which.min(result$mspe)]
  abline(v=log(1+lambda.min),col=2,lty=2)
  plot(mspe~df, result, col=3)
  df.min <- result$df[which.min(result$mspe)]
  abline(v=df.min,col=3,lty=2)
  return(result)
}

```



## Ridge regression based on $MSPE_{k-CV}(\lambda)$
Now, we will write an R function implementing the ridge regression penalization parameter$\lambda$ choice based on k-fold cross-validation $MSPE_{k−CV}(\lambda)$. 

```{r}

ridge_cv <- function(x, y, lambda.v, cv=10) {
  result <- data.frame(lambda=lambda.v, mspe=rep(0,length(lambda.v)), 
                       df=rep(0,length(lambda.v)))
  for(i in 1:length(lambda.v)) {
    lambda.v[i] <- lambda.v[i]
    folds <- createFolds(1:nrow(x), k = cv)
    result.cv <- data.frame(mspe=rep(0,cv), df=rep(0,cv))
    for(j in 1:cv) {
      fold <- folds[[j]]
      x.train <- x[-fold,]
      y.train <- y[-fold]
      x.val <- x[fold,]
      y.val <- y[fold]
      x.svd <- svd(x.train)
      d <- x.svd$d
      v <- x.svd$v
      d_inv <- diag(1/(d*d - lambda.v[i]))
      xx_inv <- t( solve( t(x.train) %*% x.train + lambda.v[i]*diag(1,ncol(x)) ))
      beta <- xx_inv %*% t(x.train) %*% y.train
      y.hat <- x.val %*% beta
      mspe <- sum((y.val - y.hat)^2) / length(y.val)
      df <- sum(d^2 / (d^2 +lambda.v[i]))
      result.cv$mspe[j] <- mspe
      result.cv$df[j] <- df
    }
    result$mspe[i] <- mean(result.cv$mspe)
    result$df[i] <- mean(result.cv$df)
    
  }
  plot(mspe~log(1+lambda), result, col=2)
  lambda.min <- result$lambda[which.min(result$mspe)]
  abline(v=log(1+lambda.min),col=2,lty=2)
  plot(mspe~df, result, col=3)
  df.min <- result$df[which.min(result$mspe)]
  abline(v=df.min,col=3,lty=2)
  
  return(result)
  
}

```

## Comparation between penalization parameters of $MSPE_{val}(\lambda)$ and $MSPE_{k-CV}(\lambda)$

```{r prostate}
ridge_loocv_gcv <- function(x, y, lambda.v) {
  l <- length(lambda.v)
  result <- data.frame(lambda=lambda.v, loocv=rep(0,l), gcv=rep(0,l),df=rep(0,l))
  n <- nrow(x)
  x.svd <- svd(x)
  d <- x.svd$d
  v <- x.svd$v
  u <- x.svd$u
  for(i in 1:l) {
    lambda <- lambda.v[i]
    d_inv <- diag(1/(d^2 - lambda))
    xx_inv <- t( solve( t(x) %*% x + lambda*diag(1,ncol(x)) ))
    beta <- (xx_inv %*% t(x)) %*% y
    y.hat <- x %*% beta
    df <- sum(d^2 / (d^2 +lambda))
    h <- x %*% xx_inv %*% t(x)
    result$loocv[i] <- sum( ( (y - y.hat)/(1 - diag(h)) )^2 ) / n
    result$gcv[i] <- sum( ( (y - y.hat)/(1 - df/n) )^2 ) / n
    result$df[i] <- df
  }
  plot(loocv~log(1+lambda), result, col=2)
  lambda.min <- result$lambda[which.min(result$loocv)]
  abline(v=log(1+lambda.min),col=2,lty=2)
  plot(loocv~df, result, col=3)
  df.min <- result$df[which.min(result$loocv)]
  abline(v=df.min,col=3,lty=2)
  plot(gcv~log(1+lambda), result, col=2)
  lambda.min <- result$lambda[which.min(result$gcv)]
  abline(v=log(1+lambda.min),col=2,lty=2)
  plot(gcv~df, result, col=3)
  df.min <- result$df[which.min(result$gcv)]
  abline(v=df.min,col=3,lty=2)
  return(result)
}
```

```{r, include=FALSE}

prostate <- fread(file="prostate_data.txt", header=TRUE)
prostate$train <- as.logical(prostate$train)
prostate$No <- NULL
prostate$train <- as.logical(prostate$train)
prostate <- as.data.frame(prostate)
data <- prostate[,1:9]
lambda.max <- 1e5
n.lambdas <- 20
lambda.v <- exp(seq(0,log(lambda.max+1),length=n.lambdas))-1
validation.ind <- prostate$train
validation <- data[!validation.ind,]
training <- data[validation.ind,]

x <- scale(training[,1:8], center = T, scale = T)
y <- scale(training[,9], center = T, scale = F)

x.val <- scale(validation[,1:8], center = T, scale = T)
y.val <- scale(validation[,9], center = T, scale = F)
```

### Validation
```{r, echo=TRUE, fig.height=6, fig.width=12}
op<-par(mfrow=c(1,2))
result.valid <- ridge(x, y, x.val, y.val, lambda.v)
```

### 5-fold and 10-fold CV
```{r, echo=TRUE, fig.height=12, fig.width=12}
x <- scale(data[,1:8], center = T, scale = T)
y <- scale(data[,9], center = T, scale = F)
op<-par(mfrow=c(2,2))
result.5.cv <- ridge_cv(x, y, cv = 10, lambda.v)
result.10.cv <- ridge_cv(x, y, cv = 10, lambda.v)
```

\newpage
### LOOCV and GCV 
```{r, echo=TRUE, fig.height=12, fig.width=12}
op<-par(mfrow=c(3,2))
result.loocv <- ridge_cv(x, y, cv = nrow(x), lambda.v) #n-fold CV
result.loocv.gcv <- ridge_loocv_gcv(x, y, lambda.v)
```

We can see from the plots that MSPE results are more solid in LOOCV and GCV than 5-fold and 10-fold. On the other hand, if we observe the validation data seems that it behaves the same way even that the curvature is different. We have to say that prostate dataset just has 97 observations and that has an impacts.
\newpage


# Ridge regression for the Boston Housing data

The Boston Housing dataset is a classical dataset which containes the values of 506 suburbs of Boston corresponding to 1978. This dataset can be found in many places but we are going to use a version with some corrections that was provided to us, which additionally includes the UTM coordinates of the geographical centers of each neighborhood. Therefore, the variables are the following:

Variable  | Description                           | Type
:--------------:|:-----------------------------------------------------------------------------------:|:--------------:
CRIM | per capita crime rate by town |  Numeric |
ZN | proportion of residential land zoned for lots over 25,000 sq.ft. |  Numeric|
INDUS | proportion of non-retail business acres per town |  Numeric|
CHAS | Charles River dummy variable (= 1 if tract bounds river; 0 otherwise) | Factor |
NOX | nitric oxides concentration (parts per 10 million) |  Numeric|
RM | average number of rooms per dwelling |  Numeric |
AGE | proportion of owner-occupied units built prior to 1940 | Numeric |
DIS | weighted distances to five Boston employment centres | Numeric |
RAD | index of accessibility to radial highways | Numeric |
TAX | full-value property-tax rate per $10,000 |  Numeric |
PTRATIO | pupil-teacher ratio by town |  Numeric |
B | 1000(Bk - 0.63)^2 where Bk is the proportion of blacks by town | Numeric  |
LSTAT | % lower status of the population | Numeric  |
MEDV | Median value of owner-occupied homes in $1000's | Numeric |


In this exercise we are going to use ridge regression on the Boston Housing dataset to fit the regression model where the response variable is $MEDV$ and the explanatory variables are the remaining 13 variables shown in the list. As we can see when loading the data, there are more variables than the ones listed ($TOWN$, $TOWNNO$, $LON$, $LAT$, $CMEDV$). We decided not to use them since the statement explicitly defines which to use.

```{r, echo=TRUE, fig.height=4, fig.width=12, message=FALSE, warning=FALSE, results = 'hide'}
load("boston.Rdata")
names(boston.c)
dataset <- boston.c[,-c(1:5)] #"TOWN"    "TOWNNO"  "TRACT"   "LON"     "LAT"
dataset$CMEDV <- NULL
```

Besides from eliminating variables, we divided the dataset between train and test. To do so, we have decided to do 3/4 training, 1/4 test.
```{r, echo=TRUE, fig.height=4, fig.width=12, message=FALSE, warning=FALSE, results = 'hide'}
trainIndex <- createDataPartition(dataset$MEDV, p =0.75, list = F)
train <- dataset[trainIndex,]
test <- dataset[-trainIndex,]
```

We perform the ridge regression with 10 fold cross-validation in order to choose the tuning parameter lambda, we take the smallest lambda with $ridge.mod\$lambda.min$ and we use this lambda to make the predictions.
```{r, echo=TRUE, fig.height=4, fig.width=12, message=FALSE, warning=FALSE, results = 'hide'}
x <- model.matrix(MEDV~., train)[,-1]
y = train$MEDV
y.test = y[-trainIndex]
#perform cross-validation to choose tuning parameter lambda
lambda.max <- 1e5  
n.lambdas <- 50
lambda.v <- exp(seq(10^-2,log(lambda.max+1),length=n.lambdas))-1  
mod <- glmnet(x, y, alpha = 0, lambda = lambda.v)
ridge.mod <- cv.glmnet(x,y,alpha= 0, lambda = lambda.v)
bestlambda <- ridge.mod$lambda.min #lambda that results in lowest cross validation error

```

```{r, echo=FALSE, fig.height=6, fig.width=12}
library(plotmo)
glmcoef<-coef(mod,bestlambda)
coef.increase<-dimnames(glmcoef[glmcoef[,1]>0,0])[[1]]
coef.decrease<-dimnames(glmcoef[glmcoef[,1]<0,0])[[1]]
allnames<-names(coef(mod)[,
            ncol(coef(mod))][order(coef(mod)[,
            ncol(coef(mod))],decreasing=TRUE)])
plot_glmnet(mod,label=TRUE,s=bestlambda)

```


```{r, echo=TRUE, message=FALSE, warning=FALSE, result='hide'}
trainContrl <- trainControl (method="repeatedcv", number=10, repeats=10)
ridge.mod10 <- train(MEDV ~ .,data = train,  trControl=trainContrl, method='glmnet', 
                                    tuneGrid=expand.grid(alpha=0,lambda=bestlambda))

ridge <- glmnet(x, y, alpha = 0, lambda = bestlambda)
normalized <- (length(train$MEDV)-1)*var(train$MEDV)
NMSE.ridge.train.error <- crossprod(predict (ridge.mod10 ) - train$MEDV) / normalized
normTest <- (length(test$MEDV)-1)*var(test$MEDV)
sse_raw <- test$MEDV - ridge$a0- data.matrix(test[,-dim(train)[2]]) %*% ridge$beta 
NMSE.ridge.test.error <- crossprod (as.matrix(sse_raw)) /normTest
```

The final model obtained has a $Df=13$ which actually is the number of coefficients we had in our model and we have a deviance of 71.07\% which has room for improvement but is not a bad result. Also the difference we have between train and test is not very big.
We can take the following conclusions from the results:
\begin{itemize}
\item We can see that as we could expect by instinct, the price of a house increases as more rooms it has. 
\item If Charles River ($CHAR$) is near, the price of the suburb is higher. 
\item $LSTAT$ implies a decrease in the house of the price.
\item $RAD$ and $MEDV$ are directly related.
\item Increase of $PTRATIO$ is associated with house price.
\item An increase in $DIS$ normally implies a decrease in $MEDV$.
\item $NOX$ puts the value down of a house.
\item $AGE$ and $INDUS$ doesn't seem to affect. 

\end{itemize}

All the aformentioned makes sense since anyone wants to commute long ways to work, nor living in a polluted area, preferably everyone wants to be near nature, etc. 


