---
title: "Local Poisson Regression"
author: "Joel Cantero Priego & Ricard Meyerhofer Parra"
date: "14/12/2019"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library("kableExtra")
library("sm")
```

\section{Introduction}
In this assignment we are going to modify the already implemented functions $h.cv.sm.binomial$ and $loglik.CV$ to obtain a bandwidth choice method for the local Poisson based on loo-CV estimation of the expected likelihood of an independent observation.  Finally, we are going to apply it to the Country Development dataset.

\section{Bandwidth Choice for Local Possion Regression}
The loo-CV of the expected log-likelihood of an independent observation can be written like this when using $h$ as bandwidth.
\[l_{CV}(h) = \frac{1}{n}\sum_{i=1}^nlog(\hat{Pr}_h^{(-i)}(Y = y_i | X = x_i)) \]
where $\hat{Pr}_h^{(-i)}(Y = y_i | X = x_i)$ is an estimation of 
\[Pr(Y = y_i | X = x_i) = e^{-\lambda_i} \frac{\lambda_i^{y_i}}{y_i!} \]
and should be estimated by maximum local likelihood 
\[\lambda_i = {\rm I\!E}(Y | X = x_i)\]


To implement this function, we had to modify the loo-CV estimation of the expected log-likelihood as aforementioned which gave us the following code:

```{r}
loglik.CV <- function(x,y,h){
    n <- length(x)
    pred <- sapply(1:n, 
                  function(i,x,y,h){
                    sm.poisson(x=x[-i],y=y[-i],h=h,eval.points=x[i],display="none")$estimate
                  }, x,y,h)
    sum = 0
    for (i in 1:n){
      sum = sum + (log((exp(-pred[i]) * ((pred[i]**y[i])/factorial(y[i])))))/n
    }
    return(-1*sum)
}
```

Now, since we have the log-likelihood method for the Poisson distribution, we can compute the best bandwidth using it:
```{r}
h.cv.sm.poisson <- function(x,y,rg.h=NULL,l.h=10,method=loglik.CV){
  cv.h <- numeric(l.h)
  i <- 0
  gr.h <- exp( seq(log(rg.h[1]), log(rg.h[2]), l=l.h))
  for (h in gr.h){
    i <- i+1
    cv.h[i] <- method(x,y,h)
  }
  return(list(h = gr.h, cv.h = cv.h, h.cv = gr.h[which.min(cv.h)]))
}
```

\section{Local Poisson regression for the Country Development Dataset}

Now we are going to consider the country development dataset which contains information on development indicators for 179 countries. The parameters that we have are the following:

Variable name         | Description                          | Values
:--------------:|:-------------------------------------:|:-------------------------------------:
iso3| Standard of country codes | String |
country_name| Official Country name |	String |
Life.expec| Total life expectancy | Integer |
Life.expec.f| Female life expectancy | Integer |
Life.expec.m| Male life expectancy	| Integer |
le.fm| Result from Life.expec.f - Life.expec.m	| Integer |
Inf.Mort.rat|  Mortality ratio | Integer |
Agric.employ.%| Percentatge of employment that agriculture covers | Integer |

As mentioned by the statement, we are going to load the dataset and we are going to create a new variable $le.fm.r$ which corresponds to the rounded value of $le.fm$ and our goal is to fit a local Poisson regression modeling $le.fm.0$ as a function of $Life.expec$.

```{r}
options(stringsAsFactors=F)
dataset <- read.csv("HDI.2017.subset.csv", sep = ";", header = T, dec=",")
dataset$le.fm.r <- round(dataset$le.fm)
```

```{r, fig.height=6, fig.width=12}
plot(dataset$Life.expec, dataset$le.fm.r,main="Distribution of life expectancy from our dataset", xlab="Life.expec", ylab = "Rounded Life.fm" )
```

We are going to see which is the bandwidth that bests fits our model, we can see in the following graph that our minimum log-likelihood is found near 3 so is the value that we are going to use.
```{r, fig.height=6, fig.width=12}
range.h <- c(1, 12)
h.CV.loglik <- h.cv.sm.poisson(dataset$Life.expec,dataset$le.fm.r, 
                               rg.h=range.h, 
                               method=loglik.CV)
```

```{r, echo=FALSE, fig.height=6, fig.width=12}
plot(h.CV.loglik$h,h.CV.loglik$cv.h)
lines(h.CV.loglik$h, h.CV.loglik$cv.h)
text(h.CV.loglik$h, h.CV.loglik$cv.h, labels=paste0("",round(h.CV.loglik$cv.h,4)), cex= 1.2)
#kable(data.frame(h.CV.loglik$h, h.CV.loglik$cv.h))
```

In the below plot, we can see that the value we are using fits the data as we can see. The selected value is the one that was the closest from 3 which was $h.cv.loglik.h =3.017$. It is not a perfect fit but the way the data is distributed does not seem it is possible to do better.
```{r, echo=FALSE, fig.height=6, fig.width=12}
chosen.h <-  h.CV.loglik$h.cv
m1 <- sm.poisson(dataset$Life.expec, dataset$le.fm.r, h=chosen.h, 
                 main="Poisson distribution with the chosen bandwidth")
```